## Introduction

The surveydatar package provides seamless integration with Flourish, allowing you to convert cross-tabulation results into interactive charts. This vignette demonstrates how to create publication-ready visualizations from survey-style analysis using the familiar mtcars dataset.

## Setting Up Your Data

First, let's prepare the mtcars data for analysis:

```{r data-prep}
# Prepare mtcars data with car model names
mtcars2 <- mtcars %>%
mutate(model = row.names(mtcars), .before = mpg)
row.names(mtcars2) <- NULL
mtcars2$model <- as.factor(mtcars2$model)

# Create some categorical variables for better cross-tabulation
mtcars2 <- mtcars2 %>%
mutate(
  mpg_category = factor(ifelse(mpg >= 20, "High MPG (≥20)", "Low MPG (<20)")),
  cyl_category = factor(ifelse(cyl >= 6, "High Cylinders (≥6)", "Low Cylinders (<6)")),
  performance = factor(case_when(
    hp >= 200 ~ "High Performance",
    hp >= 100 ~ "Medium Performance",
    TRUE ~ "Low Performance"
  ))
)

head(mtcars2)
```

## Basic Flourish Integration

### Simple Cross-Tabulation

Start with a basic cross-tabulation and convert it to a Flourish chart:

```{r basic-flourish}
# Create a cross-tabulation
tab_result <- tab(mtcars2, cyl_category, mpg_category)

# Convert to Flourish format
fltab <- tab_to_flourish(tab_result)

# Preview the structure
print(fltab)
```

### Adding Custom Settings

Enhance your chart with custom Flourish settings:

```{r custom-settings}
# Create chart with custom styling
fltab_styled <- tab_to_flourish(
tab_result,
settings = list(
  "layout.title" = "Fuel Efficiency by Engine Size",
  "layout.subtitle" = "Analysis of mtcars dataset",
  "labels" = TRUE,
  "number_format.suffix" = "%",
  "number_format.n_dec" = 1
)
)

# Preview in browser (requires Flourish API key)
preview_flourish(fltab_styled)
```

## Advanced Examples

### Using rows_list for Complex Groupings

Create more sophisticated analyses using `rows_list()`:

```{r rows-list}
# Complex row groupings
tab_advanced <- tab(
mtcars2,
rows = rows_list(
  "All Cars" = cyl_category,
  "High Performance" = cyl_category * (hp >= 200),
  "Economy Focus" = cyl_category * (mpg >= 25)
),
cols = mpg_category
)

# Convert with appropriate chart type
fltab_advanced <- tab_to_flourish(
tab_advanced,
chart_type = "column_grouped",
settings = list(
  "layout.title" = "Car Categories by Performance and Efficiency"
)
)
```

### Mean Statistics for Numeric Analysis

Analyze continuous variables using mean statistics:

TO DO: this doesn't work currently

```{r mean-analysis}
# Average horsepower by categories
tab_hp_mean <- tab(
mtcars2,
performance,
cyl_category,
statistic = "mean",
values = "hp"
)

# Create chart optimized for mean values
fltab_hp <- tab_to_flourish(
tab_hp_mean,
chart_type = "column_grouped",
settings = list(
  "layout.title" = "Average Horsepower by Performance Category",
  "number_format.n_dec" = 0,
  "number_format.suffix" = " hp"
)
)
```

### Time Series Style Analysis

Create line charts for pseudo-time series data:

```{r line-chart}
# Create data with multiple numeric columns
mpg_by_cyl <- mtcars2 %>%
group_by(cyl) %>%
summarise(
  `1970s` = mean(mpg[model %in% c("Mazda RX4", "Datsun 710")], na.rm = TRUE),
  `1980s` = mean(mpg[!model %in% c("Mazda RX4", "Datsun 710")], na.rm = TRUE),
  .groups = "drop"
)

# Manual tab_result creation for demonstration
# (In practice, you'd structure your data differently)
tab_line <- tab(mtcars2, cyl_category, mpg_category, statistic = "mean", values = "mpg")

fltab_line <- tab_to_flourish(
tab_line,
chart_type = "line",
settings = list(
  "layout.title" = "Fuel Efficiency Trends",
  "number_format.suffix" = " mpg"
)
)
```

## Chart Type Selection

### Automatic Chart Type Detection

The package automatically selects appropriate chart types based on your data:

```{r auto-chart-types}
# Different data structures get different chart types

# 2x2 table → grouped columns
tab_2x2 <- tab(mtcars2, cyl_category, mpg_category)
fltab_2x2 <- tab_to_flourish(tab_2x2)
cat("2x2 table chart type:", fltab_2x2$chart_type, "\n")

# Single column → simple columns
tab_single <- tab(mtcars2, performance)
fltab_single <- tab_to_flourish(tab_single)
cat("Single column chart type:", fltab_single$chart_type, "\n")

# Many categories → table
tab_detailed <- tab(mtcars2, model, cyl_category)
fltab_detailed <- tab_to_flourish(tab_detailed)
cat("Detailed table chart type:", fltab_detailed$chart_type, "\n")
```

### Explicit Chart Type Control

Override automatic selection when needed:

```{r explicit-chart-types}
# Force different chart types
chart_types <- c("column_grouped", "bar_grouped", "pie", "table")

for (chart_type in chart_types) {
fltab <- tab_to_flourish(tab_result, chart_type = chart_type)
cat("Chart type:", chart_type, "- Data dimensions:",
    nrow(fltab$data), "x", ncol(fltab$data), "\n")
}
```

## Data Formatting Options

### Controlling Summary Rows and Columns

Customize which elements appear in your charts:

```{r formatting-options}
# Remove base counts and summary rows for cleaner charts
fltab_clean <- tab_to_flourish(
tab_result,
strip_base = TRUE,
strip_summary_rows = TRUE,
strip_summary_cols = TRUE
)

# Keep everything for detailed analysis
fltab_detailed <- tab_to_flourish(
tab_result,
strip_base = FALSE,
strip_summary_rows = FALSE,
strip_summary_cols = FALSE
)
```

### Percent Scaling

Control how percentages are handled:

```{r percent-scaling}
# Different scaling approaches
tab_pct <- tab(mtcars2, cyl_category, mpg_category, statistic = "column_pct")

# Auto-detect scaling (recommended)
fltab_auto <- tab_to_flourish(tab_pct, percent_scale = "auto")

# Force 0-100 scale
fltab_100 <- tab_to_flourish(tab_pct, percent_scale = "0-100")

# Force 0-1 scale
# TO DO: this needs fixing
fltab_01 <- tab_to_flourish(tab_pct, percent_scale = "0-1")
```

## Integration Workflow

### Complete Workflow Example

Here's a complete example combining everything:

```{r complete-workflow}
# 1. Create comprehensive analysis
comprehensive_tab <- tab(
  mtcars2,
  rows = rows_list(
    "All Vehicles" = cyl_category,
    "Sports Cars" = cyl_category * (hp >= 150) * (qsec <= 16),
    "Economy Cars" = cyl_category * (mpg >= 25)
  ),
  cols = mpg_category,
  statistic = "column_pct"
)

# 2. Convert with full customization
fltab_final <- tab_to_flourish(
  comprehensive_tab,
  chart_type = "column_grouped",
  strip_base = TRUE,
  settings = list(
    "layout.title" = "Vehicle Analysis: Efficiency vs Performance",
    "layout.subtitle" = "Percentage distribution across categories",
    "layout.source" = "Data: mtcars dataset",
    "labels" = TRUE,
    "number_format.suffix" = "%",
    "number_format.n_dec" = 1,
    "y.linear_max" = 100,
    "y.linear_min" = 0
  )
)

# 3. Preview (requires API key setup)
preview_flourish(fltab_final)
```

### Direct Flourish Manipulation

For advanced users, you can also manipulate Flourish objects directly:

```{r direct-flourish, eval=FALSE}
# Equivalent manual approach
library(flourishcharts)

f <- flourish(
  chart_type = fltab_final$chart_type,
  api_key = Sys.getenv("FLOURISH_API_KEY")
) %>%
  bind_line_bar_pie_data(
    data = fltab_final$data,
    label = fltab_final$bindings$label,
    value = fltab_final$bindings$value
  )

# Apply settings manually
f$x$state$layout$title <- "Vehicle Analysis: Efficiency vs Performance"
f$x$state$labels <- TRUE
f$x$state$number_format$suffix <- "%"

# Display
f
```

## Setup Requirements

To use the Flourish integration, you need:

1.  **Flourish API Key**: Sign up at [flourish.studio](https://flourish.studio) and get your API key
2.  **Environment Variable**: Set `FLOURISH_API_KEY` in your `.Renviron` file:

``` r
# Add to ~/.Renviron
FLOURISH_API_KEY=your_api_key_here
```

3.  **Required Packages**: Install `flourishcharts` and `tidyr`:

``` r
install.packages(c("flourishcharts", "tidyr"))
```

## Conclusion

The surveydatar-Flourish integration provides a powerful workflow for creating interactive visualizations from survey analysis. Key benefits include:

-   **Seamless Integration**: Direct conversion from `tab()` results to Flourish charts
-   **Intelligent Defaults**: Automatic chart type selection and formatting
-   **Full Customization**: Complete control over Flourish settings and appearance
-   **Multiple Chart Types**: Support for columns, bars, lines, pies, and tables
-   **Professional Output**: Publication-ready interactive charts

This integration bridges the gap between statistical analysis and data visualization, making it easy to create compelling presentations of your survey findings.
